Synthesizing Verified Components for Cyber Assured Systems Engineering \thanks{DISTRIBUTION STATEMENT A. Approved for public release.}

### 2023-02-23 paper/paper.tex
Revise section 4

\newcommand{\figref}[1]{Fig.~\ref{#1}}
\newcommand{\secref}[1]{Sec.~\ref{#1}}
\newcommand{\lineref}[1]{Line~\ref{#1}}
\newcommand{\linesref}[2]{Lines~\ref{#1}--\ref{#2}}

@@

The contributions of this paper relate to the aspects of \brfcs\ concerning evidence that such high-assurance components meet their intended purpose and are properly implemented.
%
We leverage the existing {\agr} model checker to provide evidence of the former and the existing verified \ckml\ compiler to provide evidence of the latter \cite{agree2013,compositional-analysis-agree,nfm:agree,cakeml}.
%
\footnote{AGREE is the acronym for \emph{Assume Guarantee Reasoning Environment}.}
%
Specifically, designers write \emph{contracts} to formally specify input and output properties of the system and its components, and \agr\ proves via model checking whether or not the system composition preserves contract requirements at component inputs and guarantees contract obligations at system outputs.
%
In this way, \agr\ provides the formal framework we use to describe the behavior of an inserted high-assurance component, such as a filter or monitor, and then prove if adding such a component to the system improves cyber-resiliency.

@@

The AGREE specification for the SW component in \secref{sec:example} is given in \figref{fig:sw}.
%
\linesref{line:sw-event-def-start}{line:sw-event-def-end} define local variables that are true when data is present on the corresponding event ports for the component.  These variables are
purely for notational convenience.
%
The contract has three assumptions:
%
\begin{compactitem}
\item
\lineref{line:sw-assume-1} and \lineref{line:sw-assume-2}:
%
when input is present it is well-formed.
%
The well-formed predicates are defined elsewhere using AGREE functions.

@@

@@

Generally, AGREE contracts do not describe the computation that a component performs.
%
This is entirely by design: AGREE is intended to be declarative and reasoning about component behavior takes place solely at the specification level.
%
It defines what a component computes and not how it computes it.
%
However, the syntax of AGREE provides enough expressiveness to support the notion of a \emph{code contract}: a contract from which an implementation can be extracted.
%
First we must discuss a class of guarantees, \emph{output guarantees}, which determine the values on all output ports of a component.

@@

\begin{definition}[Code contract]
A leaf component of the form $(I,O,A,P,\emptyset)$ is a \emph{code contract} if $\mathit{Eqs} \cup G \subseteq P$, where \[\mathit{Eqs} = \set{v_1 = e_1, \cdots , v_n = e_n} \] is a non-empty set of \konst{eq} statements and $G$ is the set of output guarantees, one for each element of $O$.
%
In the interpretation as code, the order of elements of $\mathit{Eqs}$ is important, and is simply taken to be the occurrence order of the \konst{eq} statements in the syntax.
%
Thus we will work with the \emph{list} of equations $\mathit{Eqs} = [v_1 = e_1; \cdots ; v_n = e_n]$.
\end{definition}

The semantics of \secref{sec:agree-semantics} supports a formal connection between the original contract, and verification results of the AGREE model checker being run on it, and code generated from specifications of leaf level cyber-components.
%
It also provides the root meaning at the base of a chain of translation steps moving from an AGREE contract to a CakeML executable.

@@

\begin{definition}[Code contract correctness]
Assume code contract $(I,O,A,\mathit{Eqs} \cup G \cup P,\emptyset)$.
%
The contract is \emph{correct}, if for all $E$ and $t$, whenever the assumptions (historically) hold in $E$ and evaluation steps from $E$ to $E'$, then the guarantees hold in $E'$:
%
\[
\sem{\konst{Hist}(A)}^E_t \land E' = \sem{\mathit{Eqs} \cdot G}^E_t \imp \sem{P}^{E'}_t
\]
%
\end{definition}

How does this notion of correctness integrate with the verification conditions generated by AGREE?
%
In \emph{contract correctness}, the model checker is proving the following property
%
\[
\konst{Hist}(A) \imp P
\]
%
and in \emph{code contract correctness} we essentially prove a Hoare triple of the form
%
\[
\set{\konst{Hist}(A)}\; (\mathit{Eqs} \cdot G) \; \set{P}
\]
%
This accords with intuition, namely that AGREE is a kind of \emph{program-free program logic};
%
by pulling out a program $(\mathit{Eqs}\cdot G)$ from a code contract we find ourselves in the domain of imperative program verification.
%
For example, at this point one could apply a verification condition generator to help break down the proof obligation.
%
In order to use the code contract result in system contract verification conditions, one has to show that the code contract result implies the system verification condition.

%%
% IMPERATIVE CODE CONTRACTS
%%

\subsubsection*{Imperative code contracts}

Stream-based evaluation allows values arbitrarily \emph{deep in the past} to be accessed by use of $\konst{pre}(-)$.
%
However, the programming languages targeted by SPLAT do not directly support streams.
%
We are developing a translation to eliminate such deep temporal accesses by the introduction of intermediate variables.
%
The essential property the translation needs to satisfy is a syntactic restriction, call it \emph{Pascal-like}, that supports easy translation to standard imperative languages.
%
The guiding intuition is that a Pascal-like {\Eqs} accesses at most one step in the past;
%
also, accesses to a past value of a variable are only possible until the variable is updated.
%
In informal terms, this means that ${\Eqs}\cdot G$ in a code contract has the following properties:
%
\begin{compactitem}
\item Followed-by only occurs at the top level of a variable definition, \eg, $v = e_1 \to e2$;
\item $\konst{pre}(e)$ is allowed only when $e$ is a variable; and
\item $\konst{pre}(v)$ is allowed only before the defining equation $v = e$ (and may occur in $e$).
\end{compactitem}
%
For example, the Fibonacci sequence $1,1,2,3,5,8,13,\ldots$ can be defined, albeit in a manner that is not Pascal-like, by
%
{\small
\begin{lstlisting}[style=agree]
  Fib = 1 -> pre(1 -> Fib + pre Fib)
\end{lstlisting}
}
%
\noindent or by (again this is not Pascal-like)
%
{\small
\begin{lstlisting}[style=agree]
  N = 0 -> 1 + pre(N)
  Fib = if N <= 1 then 1 else pre(Fib) + pre(pre(Fib))
\end{lstlisting}
}
%
\noindent or by (this time Pascal-like):
%
{\small
\begin{lstlisting}[style=agree]
  N = 0 -> 1 + pre(N)
  F2 = 42 -> pre(F1)
  F1 = 42 -> pre(Fib)
  Fib = if N <= 1 then 1 else F1 + F2
\end{lstlisting}
}
%
Please note that the initial values of \verb+F1+ and \verb+F2+ are never accessed, so 42 could be replaced by any number in their defining equations.
%
A Pascal-like ${\Eqs} \cdot G$ is straightforward to implement in conventional programming languages, as we will see in \secref{sec:synthesis}.

\subsection{Code contracts for the hardened system}
\label{sec:ha-contracts}

@@

The monitor informally behaves as follows: if the \emph{policy} is obeyed in the current step, and the alert has not yet been raised, then the \texttt{Response} input is sent out on \texttt{Output}, and the \texttt{Alert} output stays low \lineref{line:mon-alert}.
%
Otherwise \texttt{Alert} goes high and stays high, and there is nothing sent henceforth on \texttt{Output}.

The \texttt{alert} variable (distinct from the \texttt{Alert} port) depends not just on the policy but also on the \texttt{is\_latched} constant, which has been set at configuration time.
%
Since \texttt{is\_latched} is \konst{true}, the alert is persistent, meaning that once the alert is raised, it is always raised;
%
otherwise, it is the complement of the policy value in the current step.

The monitor policy is defined by marking, with variable \texttt{isPending}, when a request is made but not yet satisfied, and by counting the number of steps (\texttt{latency}) between requests.
%
See \linesref{line:mon-pre-pending}{line:mon-latency}:
%
The \texttt{isPending} variable is true whenever there is a request that does not coincide with a response, \ie, \texttt{isPending} holds when \emph{there is no response since the most recent request without a corresponding response}.
%
The \texttt{latency} variable starts at zero, resets on every request, and otherwise increments by one.  Note that the \texttt{preIsPending} variable is a temporary needed to make the contract Pascal-like;
%
otherwise, the current and previous value of \texttt{isPending} would be simulataneously accessed in the definition of \texttt{policy}.

The policy (\lineref{line:mon-policy}) makes sure the latency is bounded whenever there is a pending request, and it ensures that a response is a result of a request.
%
The initial step differentiates from later steps in only needing to be sure a response is a result of a request, both happening in the initial step.

@@

The policy definition only works if there is never more than one outstanding request at a time;
%
otherwise, the latency counter resets incorrectly.
%
That requirement is encoded in the assumption (\lineref{line:mon-assume}) and is the same assumption used for the system input.
%
In fact, the policy can be written without the assumption, but the policy is much simpler to write with the assumption.

As shown in \figref{fig:hardened-certificate}, the cyber-hardened system passes AGREE verification.
%
That means that the high-assurance components protect the WM from the untrusted AI generating malformed data, spontaneous responses, or delayed responses.
%
Synthesis must generate faithful implementations of the code contracts for the AGREE verification result to apply to the deployed system.

@@

\section{Synthesis to CakeML}
\label{sec:synthesis}

Synthesis maps from model and specifications to code.
%
The \splt\ tool traverses the system architecture looking for occurrences of high-assurance components specified by code contracts;
%
for each such occurrence it generates a CakeML program.
%
\splt\ supports three kinds of behavioral specification from which to generate code.
%
In increasing order of expressiveness, these are: regular expressions, contiguity types, and code contracts.
%
In earlier work we have described how regular expressions and contiguity types are used as specification languages for verified filter implementations.
%
In this paper, we extend this \emph{specification oriented} approach to the Turing-complete specifications of code contracts in \agr.
%
We give a short review of that previous work before going into detail on the synthesis.

\paragraph{Regular expressions:\/} often the well-formedness of data on connections can be enforced by matching against a regular expression.
%
In other words, the data can be characterized by a \emph{regular language}.
%
As is well known, regular expressions can be translated to efficient deterministic finite state machine (DFA) implementations.
%
In \cite{formal-filter-synth-langsec,case-verified-filter} we report on the creation and application of table-based DFA filters from a verified regular expression-to-DFA compiler.
%
The latter paper offers a fully-worked out example synthesizing a regular expression filter and formally proving that it meets an non-terminating behavioral specification.

@@

When full computational power is needed, we can use the expressive power of AGREE code contracts to write arbitrarily complex filter or monitor operations, as we have already seen.
%
Our formal model of AGREE syntax and semantics is applied to extract code contracts and map them to equivalent logic functions.
%
In particular, given a code contract with functionality ${\Eqs}\cdot G$, input ports $\mathbb{I}$, state variables $\mathbb{S}$, and output ports $\mathbb{O}$, the semantics given in \secref{sec:code-contracts} can be applied to define a stream-oriented \emph{step} function $\konst{strmStep} : \mathbb{N}_0 \to \mathit{Env} \to \mathit{Env}$ defined as
%
\[
 \konst{strmStep}\;t \; E = \sem{\mathit{Eqs} \cdot G}^E_t
\]
%
Now suppose ${\Eqs}\cdot G$ is \emph{Pascalish}, \ie, sequentially imperative.
%
Then we can direcctly create a logic function
%
\[
\konst{stateStep} : \mathit{input} \times \mathit{state} \to \mathit{state} \times \mathit{output}
\]
%
corresponding to \konst{strmStep} that takes input and state as parameters and returns an updated state and output.
%
Thus the environment $\mathit{Env}$ holding all previous values of each variable can be replaced by \emph{parameters}, each of which is able to hold only one data element.
%
Assuming $\mathit{E}' = \konst{strmStep}\;t\; \mathit{E}$, we can formally prove that
%
\[
\vdash (\mathit{E}' \vert_{\mathbb{S}},
 \mathit{E}' \vert_{\mathbb{O}}) =
\konst{stateStep} (\mathit{E}\vert_{\mathbb{I}(t)},
                   \mathit{E}\vert_{\mathbb{S}(t)})
\]
%
In other words, projecting out the state and output components from the result of $\konst{strmStep}\;t\; \mathit{E}$ equals the result of applying \konst{stateStep} on the input and state components of $\mathit{E}$ at time $t$.
%
This theorem connects the stream semantics of the code contract to a pure logic function over the values appearing in the stream.
%
From this point we can use tools in the \ckml\ eco-system to jump from the setting of higher order logic to \ckml\ abstract syntax \cite{cakeml-translator}.
%
In this way we have moved from stream semantics to logic function to CakeML evaluation semantics in a property-preserving way.
%
Once in \ckml, further necessary features can be added, namely imperative variables for storing state between component invocations, also invocations of the (verified) foreign function interface of \ckml\ \cite{cakeml-monadic}.
%
Finally, the CakeML compiler can be applied to the program.
%
It delivers an executable which can be shown to preserve the original contract.

@@

\subsection{Generating \ckml}
For any of the three specification languages, we take the following general steps to arrive at an executable.
%
\begin{compactenum}
\item Map from specification to computable function in logic, producing a proof of equivalence.
\item Add parsers and printers (specified by contiguity types) for input and output.
\item Map to CakeML and add calls to input/output library routines via the CakeML foreign function interface.
\item Invoke the CakeML compiler.
\end{compactenum}
%
In the following, we examine further details in both filter and monitor synthesis.

@@

A filter is intended to be simple, although it may make deep semantic checks.
%
A filter has one input port and one output; messages on the input that the filter policy admits pass unchanged to the output port;
%
all others are dropped (not passed on).
%
We have investigated two kinds of filter.
%
In the first, a relatively shallow scan of the input suffices to enforce the policy.
%
For example, we have used the expressive power of regular expressions and Contiguity Types \cite{contiguity-types} to enforce \emph{lightweight} bounds constraints on GPS coordinates in UxAS messages.
%
On the other hand, a filter may need to parse the input buffer into a data structure specified in AGREE and apply a user-defined \emph{wellformedness} property, also specified in AGREE, to the data.
%
Arbitrarily complex wellformedness checks can be made in this way.
%
\figref{fig:filter-spec} shows a combination where the checking specified by {\small\verb+WELL_FORMED_AUTOMATION_RESPONSE+} depends on an underlying check specified by the contiguity type checking bounds on waypoints.

@@

\begin{figure}
  \begin{center}
    \begin{tabular}{c}
      \resizebox{\columnwidth}{!}{\usebox{\cml}}
    \end{tabular}
  \end{center}
  \caption{Synthesized CakeML for the filter.}
  \label{fig:filter-cakeml}
\end{figure}

@@

\begin{figure}
  \begin{center}
    \begin{tabular}{c}
      \resizebox{\columnwidth}{!}{\usebox{\monFn}}
    \end{tabular}
  \end{center}
  \caption{Synthesized CakeML for the monitor.}
  \label{fig:monitor-cakeml}
\end{figure}

@@

Within the software model, we have formalized some of the high-level requirements as assume-guarantee contracts.
%
We perform a formal analysis with \agr\ to verify that the model satisfies the contracts.
%
For the initial version of our design, the verification passes.
%
Although we are satisfied with the results of the formal verification using AGREE, we have not yet analyzed the design for cyber-vulnerabilities.

>>>>

\newcommand{\figref}[1]{Fig.~\ref{#1}}
\newcommand{\secref}[1]{Sec.~\ref{#1}}
\newcommand{\lineref}[1]{Line~\ref{#1}}
\newcommand{\defref}[1]{Def.~\ref{#1}}
\newcommand{\linesref}[2]{Lines~\ref{#1}--\ref{#2}}

@@

The contributions of this paper relate to the aspects of \brfcs\ concerning evidence that such high-assurance components meet their intended purpose and are properly implemented.
%
We leverage the existing {\agr} model checker to provide evidence of the former and the existing verified \ckml\ compiler to provide evidence of the latter \cite{agree2013,compositional-analysis-agree,nfm:agree,cakeml}.
%
\footnote{\agr\ is the acronym for \emph{Assume Guarantee Reasoning Environment}.}
%
Specifically, designers write \emph{contracts} to formally specify input and output properties of the system and its components, and \agr\ proves via model checking whether or not the system composition preserves contract requirements at component inputs and guarantees contract obligations at system outputs.
%
In this way, \agr\ provides the formal framework we use to describe the behavior of an inserted high-assurance component, such as a filter or monitor, and then prove if adding such a component to the system improves cyber-resiliency.

@@

The \agr\ specification for the SW component in \secref{sec:example} is given in \figref{fig:sw}.
%
\linesref{line:sw-event-def-start}{line:sw-event-def-end} define local variables that are true when data is present on the corresponding event ports for the component.  These variables are
purely for notational convenience.
%
The contract has three assumptions:
%
\begin{compactitem}
\item
\lineref{line:sw-assume-1} and \lineref{line:sw-assume-2}:
%
when input is present it is well-formed.
%
The well-formed predicates are defined elsewhere using \agr\ functions.

@@

%%
% Code contracts for High-assurance Components
%%

@@

\agr\ contracts do not describe the computation that a component performs by design:
%
it is intended to be declarative so it reasons about component behavior at the specification level.
%
It defines what a component computes and not how it computes it.
%
However, the syntax of \agr\ provides enough expressiveness to support the notion of a \emph{code contract}: a contract from which an implementation can be extracted.
%
First we must discuss a class of guarantees, \emph{output guarantees}, which determine the values on all output ports of a component.

@@

\begin{definition}[Code contract]
A leaf component of the form $(I,O,A,P,\emptyset)$ is a \emph{code contract} if $(\mathit{Eqs} \cup G) \subseteq P$, where \[\mathit{Eqs} = [v_1 = e_1, \cdots , v_n = e_n] \] is a non-empty list of \konst{eq} statements and $G$ is the set of output guarantees, one for each element of $O$.
%
In the interpretation as code, the order of elements of $\mathit{Eqs}$ is important and is simply taken to be the occurrence order of the \konst{eq} statements in the syntax.
\end{definition}

The semantics of \secref{sec:agree-semantics} give a formal connection between the system contract, the verification results of the \agr\ model checker, and the code synthesized for leaf level components.
%
It thus provides the root meaning at the base of a chain of translation steps moving from a code contract to a \ckml\ executable.

@@

\begin{definition}[Code contract correctness]
Assume code contract $(I,O,A,\mathit{Eqs} \cup G \cup P,\emptyset)$.
%
The contract is \emph{correct}, if for all $E$ and $t$, whenever the assumptions (historically) hold in $E$ and evaluation steps from $E$ to $E'$, then the guarantees hold in $E'$:
%
\[
(\sem{\konst{Hist}(A)}^E_t \land E' = \sem{\mathit{Eqs} \cdot G}^E_t) \imp \sem{P}^{E'}_t
\]
%
\end{definition}\label{def:correct}

How does this notion of correctness integrate with the verification conditions generated by \agr?
%
In \agr, the model checker is proving the following property
%
\[
\konst{Hist}(A) \imp P
\]
%
and in \defref{def:correct} we essentially prove a Hoare triple of the form
%
\[
\set{\konst{Hist}(A)}\; (\mathit{Eqs} \cdot G) \; \set{P}
\]
%
This accords with intuition, namely that \agr\ is a kind of \emph{program-free program logic};
%
by pulling out a program $(\mathit{Eqs}\cdot G)$ from a code contract we find ourselves in the domain of imperative program verification.
%
For example, at this point one could apply a verification condition generator to help break down the proof obligation.
%
In order to use a code contract's computed result in verification, we have to show it's defined computation implies the system verification condition as stated by \defref{def:correct}.

\paragraph{Rewriting code contracts to be imperative:\/} stream evaluation allows values arbitrarily deep in the past to be accessed by use of $\konst{pre}(-)$, but streams, and accesses to past variable values, are not part of \ckml\ and most other programming languages.
%
As such, we automatically rewrite code contracts to eliminate deep temporal accesses by the introduction of intermediate variables.
%
The essential property of the rewrite is a syntactic restriction that supports easy translation to standard imperative languages.
%
The guiding intuition is that accesses to values by expressions in {\Eqs} are at most one step in the past and accesses to such past values are only possible until the associated variable is updated.
%
In informal terms, this means that ${\Eqs} \cdot G$ in a code contract has the following properties:
%
\begin{compactitem}
\item the followed-by operator only occurs at the top level of a variable definition, \eg, $v = e_1 \to e_2$;
\item $\konst{pre}(e)$ is allowed only when $e$ is a variable; and
\item $\konst{pre}(v)$ is allowed only before the defining equation $v = e$ (and may occur in $e$).
\end{compactitem}

For example, the Fibonacci sequence is computed by
%
{\small
\begin{lstlisting}[style=agree]
  Fib = 1 -> pre(1 -> Fib + pre Fib)
\end{lstlisting}
}
%
\noindent and with a temporary variable can be rewritten to
%
{\small
\begin{lstlisting}[style=agree]
  N = 0 -> 1 + pre(N)
  Fib = if N <= 1 then 1
         else pre(Fib) + pre(pre(Fib))
\end{lstlisting}
}
%
\noindent and finally, with a few more temporary variables, can be rewritten to
%
{\small
\begin{lstlisting}[style=agree]
  N = 0 -> 1 + pre(N)
  F2 = 42 -> pre(F1)
  F1 = 42 -> pre(Fib)
  Fib = if N <= 1 then 1 else F1 + F2
\end{lstlisting}
}
%
\noindent This final version of the computation is imperative as the instances of $\konst{pre}$ can be resolved with the current values of the referenced variables:
%
the instances appear before the associated variables are updated when following a sequential order of evaluation.
%
Please note that the initial values of \verb+F1+ and \verb+F2+ are never accessed: 42 could be replaced by any number in their defining equations.
%
An imperative ${\Eqs} \cdot G$ is straightforward to implement in conventional programming languages, as we will see in \secref{sec:synthesis}.

\subsection{Code Contracts for the Hardened System}
\label{sec:ha-contracts}

@@

The monitor informally behaves as follows: if the \emph{policy} is obeyed in the current step, and the alert has not yet been raised, then the \texttt{Response} input is sent out on \texttt{Output}, and the \texttt{Alert} output stays low \lineref{line:mon-alert};
%
otherwise, \texttt{Alert} goes high and stays high, and there is nothing sent henceforth on \texttt{Output}.

The \texttt{alert} variable (distinct from the \texttt{Alert} port) depends not just on the policy but also on the constant \texttt{is\_latched}, which has been set at configuration time.
%
Since \texttt{is\_latched} is \konst{true}, the alert is persistent, meaning that once the alert is raised, it is always raised;
%
otherwise, it is the complement of the policy value in the current step.

The monitor policy is defined by marking, with variable \texttt{isPending}, when a request is made but not yet satisfied, and by counting the number of steps (\texttt{latency}) between requests.
%
See \linesref{line:mon-pre-pending}{line:mon-latency}:
%
The \texttt{isPending} variable is true whenever there is a request that does not coincide with a response, \ie, \texttt{isPending} holds when there is no response since the most recent request without a corresponding response.
%
The \texttt{latency} variable starts at zero, resets on every request, and otherwise increments by one.  Note that the \texttt{preIsPending} variable is a temporary that makes the contract imperative;
%
otherwise, the current and previous value of \texttt{isPending} would be simultaneously accessed in the definition of \texttt{policy}.

The policy (\lineref{line:mon-policy}) ensures the latency is bounded whenever there is a pending request and that a response is a result of a request.
%
The initial step differentiates from later steps in only needing to be sure that a response is a result of a request with both happening in that initial step.

@@

The policy definition only works if there is never more than one outstanding request at a time;
%
otherwise, the latency counter resets incorrectly.
%
That requirement is encoded in the assumption (\lineref{line:mon-assume}) and is the same assumption used for the system input.
%
Although the policy can be written without the assumption, it is much simpler to write with the assumption.

As shown in \figref{fig:hardened-certificate}, the cyber-hardened system passes AGREE verification.
%
That means that the high-assurance components protect the WM from the untrusted AI generating malformed data, spontaneous responses, or delayed responses.
%
Synthesis must generate faithful implementations of the code contracts for the \agr\ verification result to apply to the deployed system.

@@

\section{Synthesis to \ckml}
\label{sec:synthesis}

Synthesis maps from model and specifications to code.
%
The \splt\ tool traverses the system architecture looking for occurrences of high-assurance components specified by code contracts;
%
for each such occurrence it generates a \ckml\ program.
%
\splt\ supports three kinds of behavioral specification from which to generate code.
%
In increasing order of expressiveness, these are: regular expressions, contiguity types, and code contracts.
%
In earlier work we have described how regular expressions and contiguity types are used as specification languages for verified filter implementations.
%
In this paper, we extend this \emph{specification oriented} approach to the Turing-complete specifications of code contracts in \agr.
%
We give a short review of that previous work before going into detail on the synthesis.

\paragraph{Regular expressions:\/} often the well-formed property of data on connections can be enforced by matching against a regular expression.
%
In other words, the data can be characterized by a \emph{regular language}.
%
As is well known, regular expressions can be translated to efficient deterministic finite state machine (DFA) implementations.
%
In \cite{formal-filter-synth-langsec,case-verified-filter} we report on the creation and application of table-based DFA filters from a verified regular expression-to-DFA compiler.
%
The latter paper offers a fully-worked out example synthesizing a regular expression filter and formally proving that it meets an non-terminating behavioral specification.

@@

When full computational power is needed, we can use the expressive power of \agr\ code contracts to write arbitrarily complex filter or monitor operations, as we have already seen.
%
Our formal model of \agr\ syntax and semantics is applied to extract code contracts and map them to equivalent logic functions.
%
In particular, given a code contract with functionality ${\Eqs}\cdot G$, input ports $\mathbb{I}$, state variables $\mathbb{S}$, and output ports $\mathbb{O}$, the semantics given in \secref{sec:code-contracts} can be applied to define a stream-oriented \emph{step} function $\konst{strmStep} : \mathbb{N}_0 \to \mathit{Env} \to \mathit{Env}$ defined as
%
\[
 \konst{strmStep}\;t \; E = \sem{\mathit{Eqs} \cdot G}^E_t
\]
%
Now suppose ${\Eqs}\cdot G$ is \emph{Pascalish}, \ie, sequentially imperative.
%
Then we can direcctly create a logic function
%
\[
\konst{stateStep} : \mathit{input} \times \mathit{state} \to \mathit{state} \times \mathit{output}
\]
%
corresponding to \konst{strmStep} that takes input and state as parameters and returns an updated state and output.
%
Thus the environment $\mathit{Env}$ holding all previous values of each variable can be replaced by \emph{parameters}, each of which is able to hold only one data element.
%
Assuming $\mathit{E}' = \konst{strmStep}\;t\; \mathit{E}$, we can formally prove that
%
\[
\vdash (\mathit{E}' \vert_{\mathbb{S}},
 \mathit{E}' \vert_{\mathbb{O}}) =
\konst{stateStep} (\mathit{E}\vert_{\mathbb{I}(t)},
                   \mathit{E}\vert_{\mathbb{S}(t)})
\]
%
In other words, projecting out the state and output components from the result of $\konst{strmStep}\;t\; \mathit{E}$ equals the result of applying \konst{stateStep} on the input and state components of $\mathit{E}$ at time $t$.
%
This theorem connects the stream semantics of the code contract to a pure logic function over the values appearing in the stream.
%
From this point we can use tools in the \ckml\ eco-system to jump from the setting of higher order logic to \ckml\ abstract syntax \cite{cakeml-translator}.
%
In this way we have moved from stream semantics to logic function to CakeML evaluation semantics in a property-preserving way.
%
Once in \ckml, further necessary features can be added, namely imperative variables for storing state between component invocations, also invocations of the (verified) foreign function interface of \ckml\ \cite{cakeml-monadic}.
%
Finally, the \ckml compiler can be applied to the program.
%
It delivers an executable which can be shown to preserve the original contract.

@@

\subsection{Generating \ckml}
For any of the three specification languages, we take the following general steps to arrive at an executable.
%
\begin{compactenum}
\item Map from specification to computable function in logic, producing a proof of equivalence.
\item Add parsers and printers (specified by contiguity types) for input and output.
\item Map to \ckml\ and add calls to input/output library routines via the \ckml\ foreign function interface.
\item Invoke the \ckml\ compiler.
\end{compactenum}
%
In the following, we examine further details in both filter and monitor synthesis.

@@

A filter is intended to be simple, although it may make deep semantic checks.
%
A filter has one input port and one output; messages on the input that the filter policy admits pass unchanged to the output port;
%
all others are dropped (not passed on).
%
We have investigated two kinds of filter.
%
In the first, a relatively shallow scan of the input suffices to enforce the policy.
%
For example, we have used the expressive power of regular expressions and Contiguity Types \cite{contiguity-types} to enforce \emph{lightweight} bounds constraints on GPS coordinates in UxAS messages.
%
On the other hand, a filter may need to parse the input buffer into a data structure specified in \agr\ and apply a user-defined \emph{wellformedness} property, also specified in \agr, to the data.
%
Arbitrarily complex wellformedness checks can be made in this way.
%
\figref{fig:filter-spec} shows a combination where the checking specified by {\small\verb+WELL_FORMED_AUTOMATION_RESPONSE+} depends on an underlying check specified by the contiguity type checking bounds on waypoints.

@@

\begin{figure}
  \begin{center}
    \begin{tabular}{c}
      \resizebox{\columnwidth}{!}{\usebox{\cml}}
    \end{tabular}
  \end{center}
  \caption{Synthesized \ckml\ for the filter.}
  \label{fig:filter-cakeml}
\end{figure}

@@

\begin{figure}
  \begin{center}
    \begin{tabular}{c}
      \resizebox{\columnwidth}{!}{\usebox{\monFn}}
    \end{tabular}
  \end{center}
  \caption{Synthesized \ckml\ for the monitor.}
  \label{fig:monitor-cakeml}
\end{figure}

@@

Within the software model, we have formalized some of the high-level requirements as assume-guarantee contracts.
%
We perform a formal analysis with \agr\ to verify that the model satisfies the contracts.
%
For the initial version of our design, the verification passes.
%
Although we are satisfied with the results of the formal verification using \agr, we have not yet analyzed the design for cyber-vulnerabilities.
